
\label{sec:bak}
RISC-V is an emerging open-source software and hardware ecosystem that has gained in popularity in both industry and academia [2, 11]. At the heart of the ecosystem, the RISC-V ISA is designed to be open, simple, extensible, and free to use. The RISC-V software tool chain includes open-source compilers (e.g., GNU/GCC and LLVM), a full Linux port, a GNU/GDB debugger, verification tools, and simulators. On the hardware side, several RISC-V prototypes (e.g., Celerity [4]) have been proposed. The rapid growth of the RISC-V ecosystem enables computer architects to quickly leverage RISC-V in their research.

RISC-V ISA has a modular design and defines parts of its ISA as extensions that are coded with a letter and written as ``RV32I'' or ``RV64IMAFD''. There are small differences in the same extension for different register sizes. If the architecture (32/64 bit) is not important it can be left out and described as ``RVI'' or ``RVIMAF''. The ISA base and its extensions are developed in a collective effort between industry, the research community and educational institutions. Extensions marked as \emph{frozen} are not expected to change in any way except clarifications and improvements in documentation.

The base set of RISC-V is the Base Integer Instruction Set ``RV32/64/128I'' or ``RV32E'' (a reduced version of RV32I that supports only 16 registers designed for embedded systems). This set by itself can implement a simplified general-purpose computer, with full software support, including a general-purpose compiler.[4]

A computer design may add further extensions: Integer multiplication and division (M), Atomic instructions (A) for handling real-time concurrency, IEEE Floating point (F) with Double-precision (D) and Quad-precision (Q) options [4]. There is also an optional \emph{compact} (C) extension to reduce code size. Many RISC-V computers might add the compact extension to reduce power consumption, code size, and memory usage [4]. There are future plans for to support hypervisors, virtualization,[18] bit-manipulation (B), decimal floating-point (L), Packed SIMD (i.e. budget multimedia, P), vector processing (V) and transactional memory (T).

A small 32-bit computer for an embedded system might use ``RV32EC''. On the other hand, a large 64-bit computer system might employ ``RV64IMAFDC''. A computer with the instruction sets ``IMAFD'', an ``RV32/ 64IMAFD'', is said to be \emph{general-purpose}, summarized as G. Together with the \emph{privileged} instruction set extension an ``RVGC'' defines all instructions needed to conveniently support a Unix-style operating system.


